---
title: "Lab 02 - Plastic waste"
author: "Eric Stone"
date: "1.8.24"
output: github_document
---

## Load packages and data

```{r load-packages, message=FALSE}
library(tidyverse) 
```

```{r load-data, message=FALSE}
plastic_waste <- read.csv("data/plastic-waste.csv")
```

## Exercises

### Exercise 1

This was interesting to play around with. You can see my various attempts at visualization below. It turned out to be most effective to take Trinidad and Tobago out. In general, Europe and North America seem to have the most waste. Africa has the least. I expect that this is mostly a function of how developed the nations are. 

```{r plastic-waste-continent}
ggplot(data = plastic_waste, aes(x = plastic_waste_per_cap)) +
  geom_histogram(binwidth = 0.2)
plastic_waste %>%
  filter(plastic_waste_per_cap > 3.5)
ggplot(data = plastic_waste, aes(x = plastic_waste_per_cap)) +
  geom_histogram(binwidth = 0.2)+
  facet_grid(~ continent)
ggplot(data = plastic_waste, aes(x = plastic_waste_per_cap)) +
  geom_histogram(binwidth = 0.8)+
  facet_grid(~ continent)
ggplot(data = plastic_waste, aes(x = plastic_waste_per_cap)) +
  geom_histogram(binwidth = 0.1)+
  facet_grid(~ continent)
ggplot(data = plastic_waste, aes(x = plastic_waste_per_cap)) +
  geom_histogram(binwidth = 0.1)+
  facet_wrap(~ continent, ncol = 2)
ggplot(data = plastic_waste %>%
  filter(plastic_waste_per_cap < 3.5), aes(x = plastic_waste_per_cap)) +
  geom_histogram(binwidth = 0.1)+
  facet_wrap(~ continent, ncol = 1) 
ggplot(data = plastic_waste %>%
  filter(plastic_waste_per_cap < 3.5), aes(x = plastic_waste_per_cap)) +
  geom_histogram(binwidth = 0.1)+
  facet_wrap(~ continent, ncol = 2) 
```

### Exercise 2

Describe why we defined the color and fill of the curves by mapping aesthetics of the plot but we defined the alpha level as a characteristic of the plotting geom. 

I believe this is a mapping vs. setting issue. "color" and "fill" were based on the continent variable, but alpha was chosen by us, not based on any particular variable. Note this isn't a function of, for example, color vs. alpha, but instead with what we were using these for in this particular case.

```{r plastic-waste-density}
ggplot(data = plastic_waste,
       aes(x = plastic_waste_per_cap)) +
  geom_density()
ggplot(data = plastic_waste, 
       mapping = aes(x = plastic_waste_per_cap, 
                     color = continent)) +
  geom_density()
ggplot(data = plastic_waste, 
       mapping = aes(x = plastic_waste_per_cap, 
                     color = continent, 
                     fill = continent)) +
  geom_density()
ggplot(data = plastic_waste, 
       mapping = aes(x = plastic_waste_per_cap, 
                     color = continent, 
                     fill = continent)) +
  geom_density(alpha = 0.7)
ggplot(data = plastic_waste, 
       mapping = aes(x = plastic_waste_per_cap, 
                     color = continent, 
                     fill = continent)) +
  geom_density(alpha = 0.2)
ggplot(data = plastic_waste, 
       mapping = aes(x = plastic_waste_per_cap, 
                     color = continent, 
                     fill = continent)) +
  geom_density(alpha = 0.05)
```

### Exercise 3

What do the violin plots reveal that box plots do not? What features are apparent in the box plots but not in the violin plots?  

At least given these defaults, box plots provide more summary statistics (like (I think) the 25% and 75% quartiles) than the violin plots do. They also seem to convey the outliers more clearly. I believe violin plots are used when one wants to have a pretty presentation, but doesn't really care about communicating information. :) 

```{r plastic-waste-box plot}
ggplot(data = plastic_waste, 
       mapping = aes(x = continent, 
                     y = plastic_waste_per_cap)) +
  geom_boxplot()
ggplot(data = plastic_waste, 
       mapping = aes(x = continent, 
                     y = plastic_waste_per_cap)) +
  geom_violin()
```


### Exercise 4

4.1. Visualize the relationship between plastic waste per capita and mismanaged plastic waste per capita using a scatterplot. Describe the relationship.

In general, as the amount of plastic waste per capita increases, the amount of mismanaged plastic waste per capita increases as well. It seems to be a moderately strong relationship.

4.2. Color the points in the scatterplot by continent. Does there seem to be any clear distinctions between continents with respect to how plastic waste per capita and mismanaged plastic waste per capita are associated?

I was having trouble seeing this with the graphs overlayed, so I faceted by continent and then eliminated T+T from the graph. The relationship seems particularly strong in Africa, Oceania, and (to a somewhat lesser extent) Asia. Interestingly, there doesn't seem to be any relationship in Europe and at most a small relationship in North America. I expect the lack of relationship in Europe is because most of the countries are developed there.

4.3. Visualize the relationship between plastic waste per capita and total population as well as plastic waste per capita and coastal population. You will need to make two separate plots. Do either of these pairs of variables appear to be more strongly linearly associated?

There seems to be a small to moderate negative relationship in both cases. Maybe it's slightly greater with coastal population, but it doesn't seem that different to me. Note to visualize what was going on, I again took T+T out. 

```{r plastic-waste-mismanaged}
ggplot(data = plastic_waste, mapping = aes(x = plastic_waste_per_cap, y = mismanaged_plastic_waste_per_cap)) +
  geom_point() +
  labs(title = "Plastic waste per capita vs. Mismanaged plastic waste per capita",
       x = "Plastic waste", y = "Mismanaged plastic waste")
ggplot(data = plastic_waste, mapping = aes(x = plastic_waste_per_cap, y = mismanaged_plastic_waste_per_cap, color = continent)) +
  geom_point() +
  labs(title = "Plastic waste per capita vs. Mismanaged plastic waste per capita, per continent",
       x = "Plastic waste", y = "Mismanaged plastic waste")
ggplot(data = plastic_waste, mapping = aes(x = plastic_waste_per_cap, y = mismanaged_plastic_waste_per_cap, color = continent)) +
  geom_point() + 
  facet_wrap(~ continent, ncol = 2) +
  labs(title = "Plastic waste per capita vs. Mismanaged plastic waste per capita, per continent",
       x = "Plastic waste", y = "Mismanaged plastic waste")
ggplot(data = plastic_waste %>%
  filter(plastic_waste_per_cap < 3.5), mapping = aes(x = plastic_waste_per_cap, y = mismanaged_plastic_waste_per_cap, color = continent)) +
  geom_point() + 
  facet_wrap(~ continent, ncol = 2) +
  labs(title = "Plastic waste per capita vs. Mismanaged plastic waste per capita, per continent",
       x = "Plastic waste", y = "Mismanaged plastic waste")
ggplot(data = plastic_waste, mapping = aes(x = plastic_waste_per_cap, y = total_pop)) +
  geom_point() +
  labs(title = "Plastic waste per capita vs. Total Population",
       x = "Plastic waste", y = "Total population")
ggplot(data = plastic_waste, mapping = aes(x = plastic_waste_per_cap, y = coastal_pop)) +
  geom_point() +
  labs(title = "Plastic waste per capita vs. Coastal population",
       x = "Plastic waste", y = "Coastal population")
ggplot(data = plastic_waste %>%
  filter(plastic_waste_per_cap < 3.5), mapping = aes(x = plastic_waste_per_cap, y = total_pop)) +
  geom_point() +
  labs(title = "Plastic waste per capita vs. Total Population",
       x = "Plastic waste", y = "Total population")
ggplot(data = plastic_waste %>%
  filter(plastic_waste_per_cap < 3.5), mapping = aes(x = plastic_waste_per_cap, y = coastal_pop)) +
  geom_point() +
  labs(title = "Plastic waste per capita vs. Coastal population",
       x = "Plastic waste", y = "Coastal population")
```

### Exercise 5

Note I had trouble specifying the dataset for the analysis. I saw the "attach" command on-line. Is there a better way to do this?

```{r plastic-waste-mismanaged-continent}
attach(plastic_waste)
perc_coastal <- coastal_pop / total_pop 
ggplot(data = plastic_waste %>%
  filter(plastic_waste_per_cap < 3.5), mapping = aes(x = perc_coastal, y = plastic_waste_per_cap)) +
  geom_point() +
  labs(title = "Coastal population percentage vs. Plastic waste per capita",
       x = "Coastal population percentage", y = "Plastic waste")
```


